Variational Approximations for Transfer in Reinforcement Learning

### 2018-05-15 NIPS 2018/main.tex
Merged with Marcello's edits

Recent advancements have allowed reinforcement learning (RL) \cite{sutton1998reinforcement} to achieve impressive results in a wide variety of complex tasks, ranging from Atari \cite{mnih2015human}, to the game of Go \cite{silver2016mastering}, to the control of sophisticated robotics systems \cite{kober2009policy,lillicrap2015continuous,levine2016end}. The main limitation is that RL algorithms still require an enormous amount of experience samples before successfully learning such complicated tasks. One of the most promising solutions is transfer learning, which focuses on reusing past knowledge available to the agent in order to reduce the sample-complexity for learning new tasks. In the typical settings of transfer in RL \cite{taylor2009transfer}, the agent is assumed to have already solved a set of \textit{source tasks} generated from some unknown distribution. Then, given a \textit{target task} drawn from the same distribution, or a slightly different one, the agent can rely on knowledge from the source tasks to speed up the learning process. This constitutes a significant advantage over plain RL, where the agent learns each new task from scratch independently of previous learning experience. Several algorithms have been proposed in the literature to transfer experience samples \cite{lazaric2008transfer,taylor2008transferring}, policies/options \cite{fernandez2006probabilistic,konidaris2007building}, rewards \cite{konidaris2006autonomous}, features \cite{barreto2017successor}, parameters \cite{doshi2016hidden,killian2017robust}, and so on. We refer the reader to \cite{taylor2009transfer} for a thorough survey on transfer in RL.

%\paragraph*{Exploration-based introduction}
%One of the most relevant problems in this context is how to efficiently explore the target task based on knowledge from the source tasks. Intuitively, assuming the tasks under consideration share some similarities due to the common distribution, much better exploration strategies than uninformed ones (e.g., $\epsilon$-greedy) can be adopted for quickly learning the target task. Among the appealing approaches for this problem we find Bayesian methods (e.g., \cite{}), which are able to model the uncertainty over the current task based on previous knowledge and drive exploration so that this uncertainty is reduced as quickly as possible. Similarly, model-based algorithms (e.g., \cite{}) typically transfer samples to improve their estimates of the task model and adopt classic count-based exploration to drive the agent towards regions where such estimates are more uncertain. However, all these approaches either require strong assumptions (for example, on the distribution involved in Bayesian methods) or do not scale well to large problems. This greatly limits their practical applicability.

%In this work, we tackle such limitations by considering a more general approach. Similarly to \cite{}, we assume tasks to share similarities in their value functions and use the given source tasks to learn the distribution over such functions. Then, we use this distribution as a prior for learning the target task and we propose and efficient variational approximation of the corresponding posterior. Leveraging on recent ideas from randomized value functions (\cite{}), we design a Thompson sampling-based algorithm which efficiently explores the target task by repeatedly sampling from the posterior and acting greedily w.r.t. (with respect to) the sampled value function. We show that our approach is very general, in the sense that it does not require any specific choice of function approximator or prior/posterior distribution models. \todo[inline]{Maybe mention something about theory/experiments here}

Under the assumption that tasks follow a certain distribution, an intuitive choice for designing a transfer algorithm is to attempt at characterizing the uncertainty over the target task. Then, an ideal algorithm would leverage prior knowledge from the source tasks to interact with the target task in such a way that this uncertainty is reduced as quickly as possible. This simple intuition makes Bayesian \todo{probabilistic? distributional?} methods appealing approaches for transfer in RL, and many previous works have been proposed in this direction. \cite{wilson2007multi} assume tasks share similarities in their dynamics and rewards and propose a hierarchical Bayesian model for the distribution of these two elements. Similarly, \cite{lazaric2010bayesian} assume tasks are similar in their value functions and design a different hierarchical Bayesian model for transferring such information. More recently, \cite{doshi2016hidden}, and its extension \cite{killian2017robust}, consider tasks whose dynamics are governed by some hidden parameters, and propose efficient Bayesian models for quickly learning such parameters in new tasks. However, all these algorithms require specific, and sometimes restrictive, assumptions (e.g., on the distributions involved or the function approximators adopted), which might limit their practical applicability. The importance of having transfer algorithms that alleviate the need of strong assumptions and that can be easily adapted to different contexts motivates us to take a more general approach.

Similarly to \cite{lazaric2010bayesian}, we assume tasks to share similarities in their value functions and use the given source tasks to learn a distribution over such functions. Then, we use this distribution as a prior for learning the target task and we propose and efficient variational approximation of the corresponding posterior. Leveraging on recent ideas from randomized value functions \cite{osband2014generalization}, we design a Thompson Sampling-based algorithm which efficiently explores the target task by repeatedly sampling from the posterior and acting greedily w.r.t. (with respect to) the sampled value function. We show that our approach is very general, in the sense that it does not require any specific choice of function approximator or prior/posterior distribution models. We propose a finite-sample analysis that theoretically validates our approach, while empirically evaluating its performance on three domains with increasing level of difficulty.

@@

We consider a distribution $\mathcal{D}$ over tasks, where each task is modeled as a discounted Markov Decision Process (MDP). We define an MDP as a tuple $\mathcal{M} = \langle \mathcal{S},\mathcal{A},\mathcal{P},\mathcal{R},p_0,\gamma\rangle$\todo{Should we provide a more transfer-oriented definition?}, where $\mathcal{S}$ is the state-space, $\mathcal{A}$ is a finite set of actions, $\mathcal{P}(\cdot | s,a)$ is the distribution of the next state $s'$ given that action $a$ is taken in state $s$, $\mathcal{R}: \mathcal{S}\times\mathcal{A}\rightarrow\mathbb{R}$ is the reward function, $p_0$ is the initial-state distribution, and $\gamma\in [0,1)$ is the discount factor. We assume the reward function to be uniformly bounded by a constant $R_{max}>0$. A deterministic policy $\pi : \mathcal{S} \rightarrow \mathcal{A}$ is a mapping from states to actions. At the beginning of each episode of interaction, the initial state $s_0$ is drawn from $p_0$. Then, the agent takes the action $a_0 = \pi(s_0)$, receives a reward $\mathcal{R}(s_0,a_0)$, transitions to the next state $s_1 \sim \mathcal{P}(\cdot | s_0,a_0)$, and the process is repeated. The goal is to find the policy maximizing the long-term return over a possibly infinite horizon: $\max_{\pi}J(\pi)\triangleq\mathbb{E}[\sum_{t=0}^{\infty} \gamma^t r_t \mid \mathcal{M},\pi]$. To this end, we define the optimal value function $Q^*(s,a)$ as the expected return obtained by taking action $a$ in state $s$ and following an optimal policy thereafter. Then, an optimal policy $\pi^*$ is a policy that is greedy with respect to the optimal value function, i.e., $\pi^*(s) = \argmax_a Q^*(s,a)$ for all states $s$. It can be shown (e.g., \cite{puterman1994markov}) that $Q^*$ is the unique fixed-point of the optimal Bellman operator $T$ defined by $TQ(s,a) = \mathcal{R}(s,a) + \gamma\mathbb{E}_{\mathcal{P}}[\max_{a'}Q(s',a')]$ for any value function $Q$. From now on, we adopt the term $Q$-function to denote any plausible value function, i.e., any function $Q : \mathcal{S}\times\mathcal{A} \rightarrow \mathbb{R}$ uniformly bounded by $\frac{R_{max}}{1-\gamma}$.

We consider a parametric family of $Q$-functions, $\mathcal{Q} = \left\{ Q_{\bm{w}} : \mathcal{S}\times\mathcal{A} \rightarrow \mathbb{R} \mid \bm{w}\in\mathbb{R}^K\right\}$, and we assume each function in $\mathcal{Q}$ to be uniformly bounded by $\frac{R_{max}}{1-\gamma}$. When learning the optimal value function, a quantity of interest is how close a given function $Q_{\bm{w}}$ is to the fixed-point of the Bellman operator. A possible measure is its Bellman error (or Bellman residual), defined by $B(\bm{w}) \triangleq TQ_{\bm{w}} - Q_{\bm{w}}$. Notice that $Q_{\bm{w}}$ is optimal if, and only if, $B(\bm{w})(s,a) = 0$ for all $s,a$. If we assume the existence of a distribution $\nu$ over $\mathcal{S}\times\mathcal{A}$, an appealing objective is to directly minimize the squared Bellman error of $Q_{\bm{w}}$ under $\nu$, denoted by $\norm{B(\bm{w})}_{\nu}^2$\todo{Should we define norms before?}. Unfortunately, it is well-known that an unbiased estimator of this quanitity requires two independent samples of the next state $s'$ for each $s,a$ (e.g., \cite{maillard2010finite}). In practice, the Bellman error is typically replaced by the TD error $b(\bm{w})$, which approximates the former using a single transition sample, $b(\bm{w}) = r + \gamma\max_{a'}Q_{\bm{w}}(s',a') - Q_{\bm{w}}(s,a)$. Finally, given a dataset $D = \langle s_i,a_i,r_i,s_i' \rangle_{i=1}^{N}$ of $N$ samples, the squared TD error is computed as $\norm{B(\bm{w})}_{D}^2 = \frac{1}{N}\sum_{i=1}^N (r_i + \gamma \max_{a'} Q_{\bm{w}}(s_i',a') - Q_{\bm{w}}(s_i,a_i))^2 = \frac{1}{N}\sum_{i=1}^N b_i(\bm{w})^2$. Whenever the distinction is clear from the context, with slight abuse of terminology, we refer to the squared Bellman error and squared TD error as Bellman error and TD error, respectively.

%\subsection{Variational Inference}
%
%When working with Bayesian approaches, the posterior distribution of hidden variables $\bm{w} \in \mathbb{R}^K$ given data $D$,
%\begin{equation}\label{eq:bayes}
%p(\bm{w} | D) = \frac{p(D | \bm{w})p(\bm{w})}{p(D)} = \frac{p(D | \bm{w})p(\bm{w})}{\int_{\bm{w}} p(D | \bm{w})p(\bm{w})},
%\end{equation}
%is typically intractable for many models of interest (e.g., when working with deep neural networks) due to difficulties in computing the integral of Eq. \eqref{eq:bayes}. The main intuition behind variational inference \cite{} \todo{CITE} is to approximate the intractable posterior $p(\bm{w} | D)$ with a simpler distribution $q_{\bm{\xi}}(\bm{w})$. The latter is chosen in a parametric family, with variational parameters $\bm{\xi}$, as the minimizer of the Kullback-Leibler (KL) divergence w.r.t. $p$:
%\begin{equation}
%\min_{\bm{\xi}} KL\left(q_{\bm{\xi}}(\bm{w})\ ||\ p(\bm{w} \mid D)\right)
%\end{equation}
%It is well-known that minimizing the KL divergence is equivalent to maximizing the so-called \textit{evidence lower bound} (ELBO), which is defined as:
%\begin{equation}
%\text{ELBO}(\bm{\xi}) = \mathbb{E}_{\bm{w} \sim q_{\bm{\xi}}}\left[ \log p(D | \bm{w}) \right] - KL\left(q_{\bm{\xi}}(\bm{w})\ ||\ p(\bm{w})\right)
%\end{equation}
%Intuitively, the best approximation is the one that maximizes the expected log-likelihood of the data, while minimizing the KL divergenge w.r.t. the prior $p(\bm{w})$.

>>>>

Recent advancements have allowed reinforcement learning (RL)~\cite{sutton1998reinforcement} to achieve impressive results in a wide variety of complex tasks, ranging from Atari~\cite{mnih2015human} through the game of Go~\cite{silver2016mastering} to the control of sophisticated robotics systems~\cite{kober2009policy,lillicrap2015continuous,levine2016end}. The main limitation is that these RL algorithms still require an enormous amount of experience samples before successfully learning such complicated tasks. One of the most promising solutions to reduce the need of samples is transfer learning, which focuses on reusing past knowledge available to the agent in order to reduce the sample-complexity for learning new tasks. In the typical settings of transfer in RL~\cite{taylor2009transfer}, the agent is assumed to have already solved a set of \textit{source tasks} generated from some unknown distribution. Then, given a \textit{target task} (which is drawn from the same distribution, or a slightly different one), the agent can rely on knowledge from the source tasks to speed up the learning process. This reuse of knowledge constitutes a significant advantage over plain RL, where the agent learns each new task from scratch independently of any previous learning experience. Several algorithms have been proposed in the literature to transfer different elements involved in the learning process: experience samples~\cite{lazaric2008transfer,taylor2008transferring}, policies/options~\cite{fernandez2006probabilistic,konidaris2007building}, rewards~\cite{konidaris2006autonomous}, features~\cite{barreto2017successor}, parameters~\cite{doshi2016hidden,killian2017robust}, and so on. We refer the reader to~\cite{taylor2009transfer,lazaric2012transfer} for a thorough survey on transfer in RL.

%\paragraph*{Exploration-based introduction}
%One of the most relevant problems in this context is how to efficiently explore the target task based on knowledge from the source tasks. Intuitively, assuming the tasks under consideration share some similarities due to the common distribution, much better exploration strategies than uninformed ones (e.g., $\epsilon$-greedy) can be adopted for quickly learning the target task. Among the appealing approaches for this problem we find Bayesian methods (e.g.,~\cite{}), which are able to model the uncertainty over the current task based on previous knowledge and drive exploration so that this uncertainty is reduced as quickly as possible. Similarly, model-based algorithms (e.g.,~\cite{}) typically transfer samples to improve their estimates of the task model and adopt classic count-based exploration to drive the agent towards regions where such estimates are more uncertain. However, all these approaches either require strong assumptions (for example, on the distribution involved in Bayesian methods) or do not scale well to large problems. This greatly limits their practical applicability.

%In this work, we tackle such limitations by considering a more general approach. Similarly to~\cite{}, we assume tasks to share similarities in their value functions and use the given source tasks to learn the distribution over such functions. Then, we use this distribution as a prior for learning the target task and we propose and efficient variational approximation of the corresponding posterior. Leveraging on recent ideas from randomized value functions (\cite{}), we design a Thompson sampling-based algorithm which efficiently explores the target task by repeatedly sampling from the posterior and acting greedily w.r.t. (with respect to) the sampled value function. We show that our approach is very general, in the sense that it does not require any specific choice of function approximator or prior/posterior distribution models. \todo[inline]{Maybe mention something about theory/experiments here}

Under the assumption that tasks follow a specific distribution, an intuitive choice for designing a transfer algorithm is to attempt at characterizing the uncertainty over the target task. Then, an ideal algorithm would leverage prior knowledge from the source tasks to interact with the target task to reduce the uncertainty as quickly as possible. This simple intuition makes Bayesian \todo{probabilistic? distributional?} methods appealing approaches for transfer in RL, and many previous works have been proposed in this direction. In~\cite{wilson2007multi}, the authors assume tasks share similarities in their dynamics and rewards and propose a hierarchical Bayesian model for the distribution of these two elements. Similarly, in~\cite{lazaric2010bayesian}, the authors assume that tasks are similar in their value functions and design a different hierarchical Bayesian model for transferring such information. More recently,~\cite{doshi2016hidden}, and its extension~\cite{killian2017robust}, consider tasks whose dynamics are governed by some hidden parameters, and propose efficient Bayesian models for quickly learning such parameters in new tasks. However, most of these algorithms require specific, and sometimes restrictive, assumptions (e.g., on the distributions involved or the function approximators adopted), which might limit their practical applicability. The importance of having transfer algorithms that alleviate the need for strong assumptions and that easily adapt to different contexts motivates us to take a more general approach.

Similarly to~\cite{lazaric2010bayesian}, we assume tasks to share similarities in their value functions and use the given source tasks to learn a distribution over such functions. Then, we use this distribution as a prior for learning the target task and we propose a variational approximation of the corresponding posterior that is computationally efficient. Leveraging on recent ideas from randomized value functions~\cite{osband2014generalization}, we design a Thompson Sampling-based algorithm which efficiently explores the target task by repeatedly sampling from the posterior and acting greedily w.r.t. (with respect to) the sampled value function. We show that our approach is very general, in the sense that it can work with any parametric function approximator and with any prior/posterior distribution models (in this paper we focus on the Gaussian and Gaussian mixture models). In addition to the algorithmic contribution, we give also a theoretical contribution by providing a finite-sample analysis of our approach and an experimental contribution showing its empirical performance on four domains with increasing level of difficulty.

@@

We consider a distribution $\mathcal{D}$ over tasks, where each task $\mathcal{M}_\tau$ is modeled as a discounted Markov Decision Process (MDP). We define an MDP as a tuple $\mathcal{M}_\tau = \langle \mathcal{S},\mathcal{A},\mathcal{P}_\tau,\mathcal{R}_\tau,p_0,\gamma\rangle$, where $\mathcal{S}$ is the state-space, $\mathcal{A}$ is a finite set of actions, $\mathcal{P}_\tau(\cdot | s,a)$ is the distribution of the next state $s'$ given that action $a$ is taken in state $s$, $\mathcal{R}_\tau: \mathcal{S}\times\mathcal{A}\rightarrow\mathbb{R}$ is the reward function, $p_0$ is the initial-state distribution, and $\gamma\in [0,1)$ is the discount factor. We assume the reward function to be uniformly bounded by a constant $R_{max}>0$. A deterministic policy $\pi : \mathcal{S} \rightarrow \mathcal{A}$ is a mapping from states to actions. At the beginning of each episode of interaction, the initial state $s_0$ is drawn from $p_0$. Then, the agent takes the action $a_0 = \pi(s_0)$, receives a reward $\mathcal{R}_\tau(s_0,a_0)$, transitions to the next state $s_1 \sim \mathcal{P}(\cdot | s_0,a_0)$, and the process is repeated. The goal is to find the policy maximizing the long-term return over a possibly infinite horizon: $\max_{\pi}J(\pi)\triangleq\mathbb{E}_{\mathcal{M}_\tau,\pi}[\sum_{t=0}^{\infty} \gamma^t \mathcal{R}_\tau(s_t,a_t)]$. To this end, we define the optimal value function of task $\mathcal{M}_\tau$, $Q_\tau^*(s,a)$, as the expected return obtained by taking action $a$ in state $s$ and following an optimal policy thereafter. Then, an optimal policy $\pi_\tau^*$ is a policy that is greedy with respect to the optimal value function, i.e., $\pi_\tau^*(s) = \argmax_a Q_\tau^*(s,a)$ for all states $s$. It can be shown (e.g.,~\cite{puterman1994markov}) that $Q_\tau^*$ is the unique fixed-point of the optimal Bellman operator $T_\tau$ defined by $T_\tau Q(s,a) = \mathcal{R}_\tau(s,a) + \gamma\mathbb{E}_{s' \sim\mathcal{P}_\tau}[\max_{a'}Q(s',a')]$ for any value function $Q$. From now on, we adopt the term $Q$-function to denote any plausible value function, i.e., any function $Q : \mathcal{S}\times\mathcal{A} \rightarrow \mathbb{R}$ uniformly bounded by $\frac{R_{max}}{1-\gamma}$. In the following, to avoid cluttering the notation, we will drop the subscript $\tau$ when there is no ambiguity.

We consider a parametric family of $Q$-functions, $\mathcal{Q} = \left\{ Q_{\bm{w}} : \mathcal{S}\times\mathcal{A} \rightarrow \mathbb{R} \mid \bm{w}\in\mathbb{R}^d\right\}$, and we assume each function in $\mathcal{Q}$ to be uniformly bounded by $\frac{R_{max}}{1-\gamma}$. When learning the optimal value function, a quantity of interest is how close a given function $Q_{\bm{w}}$ is to the fixed-point of the Bellman operator. A possible measure is its Bellman error (or Bellman residual), defined by $B_{\bm{w}} \triangleq TQ_{\bm{w}} - Q_{\bm{w}}$. Notice that $Q_{\bm{w}}$ is optimal if and only if $B_{\bm{w}}(s,a) = 0$ for all $s,a$. If we assume the existence of a distribution $\nu$ over $\mathcal{S}\times\mathcal{A}$, a sound objective is to directly minimize the squared Bellman error of $Q_{\bm{w}}$ under $\nu$, denoted by $\norm{B_{\bm{w}}}_{\nu}^2$. Unfortunately, it is well-known that an unbiased estimator of this quantity requires two independent samples of the next state $s'$ for each $s,a$ (e.g.,~\cite{maillard2010finite}). In practice, the Bellman error is typically replaced by the TD error $b(\bm{w})$, which approximates the former using a single transition sample $\langle s,a,s',r \rangle$, $b(\bm{w}) = r + \gamma\max_{a'}Q_{\bm{w}}(s',a') - Q_{\bm{w}}(s,a)$. Finally, given a dataset $D = \langle s_i,a_i,r_i,s_i' \rangle_{i=1}^{N}$ of $N$ samples, the squared TD error is computed as $\norm{B_{\bm{w}}}_{D}^2 = \frac{1}{N}\sum_{i=1}^N (r_i + \gamma \max_{a'} Q_{\bm{w}}(s_i',a') - Q_{\bm{w}}(s_i,a_i))^2 = \frac{1}{N}\sum_{i=1}^N b_i(\bm{w})^2$. Whenever the distinction is clear from the context, with slight abuse of terminology, we refer to the squared Bellman error and squared TD error as Bellman error and TD error, respectively.

%\subsection{Variational Inference}
%
%When working with Bayesian approaches, the posterior distribution of hidden variables $\bm{w} \in \mathbb{R}^K$ given data $D$,
%\begin{equation}\label{eq:bayes}
%p(\bm{w} | D) = \frac{p(D | \bm{w})p(\bm{w})}{p(D)} = \frac{p(D | \bm{w})p(\bm{w})}{\int_{\bm{w}} p(D | \bm{w})p(\bm{w})},
%\end{equation}
%is typically intractable for many models of interest (e.g., when working with deep neural networks) due to difficulties in computing the integral of Eq. \eqref{eq:bayes}. The main intuition behind variational inference~\cite{} \todo{CITE} is to approximate the intractable posterior $p(\bm{w} | D)$ with a simpler distribution $q_{\bm{\xi}}(\bm{w})$. The latter is chosen in a parametric family, with variational parameters $\bm{\xi}$, as the minimizer of the Kullback-Leibler (KL) divergence w.r.t. $p$:
%\begin{equation}
%\min_{\bm{\xi}} KL\left(q_{\bm{\xi}}(\bm{w})\ ||\ p(\bm{w} \mid D)\right)
%\end{equation}
%It is well-known that minimizing the KL divergence is equivalent to maximizing the so-called \textit{evidence lower bound} (ELBO), which is defined as:
%\begin{equation}
%\text{ELBO}(\bm{\xi}) = \mathbb{E}_{\bm{w} \sim q_{\bm{\xi}}}\left[ \log p(D | \bm{w}) \right] - KL\left(q_{\bm{\xi}}(\bm{w})\ ||\ p(\bm{w})\right)
%\end{equation}
%Intuitively, the best approximation is the one that maximizes the expected log-likelihood of the data, while minimizing the KL divergenge w.r.t. the prior $p(\bm{w})$.
